Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-05-02 Manuscript/manuscript.Rmd
Results run-through

Cell type-specific genomic annotations were used to build the predictor space for our TAD boundary prediction framework. Bins were annotated by the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average [??? Is that correct? Or, max?] signal strength of the corresponding annotation (*Peak Signal Strength*), and the $log_{2}$ of the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*).

@@

We developed a computational framework<!--, refered to as *TADRF*,--> aimed at determining the optimal set of data level characteristics necessary for topologically associated domain (TAD) boundary prediction using random forest classification models. We then leveraged our optimally built random forest to precisely predict the genomic coordinates of both TAD and loop boundaries at base pair resolution.

Our framework is resolution-specific and uses arbitrary called TAD and loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries we opted to use boundaries obtained by the PEAKACHU tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised].

Boundary regions were defined as regions on the linear genome where a genomic bin contained a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a boundary ($Y = 0$) (Figure 1A). Under this construction, the number of non-boundary regions far exceeded the number of boundary regions. To cope with the imbalance among the classes of genomic bins, we evaluated three re-sampling techniques (in addition to no re-sampling) before predicting boundary regions. *Random over-sampling* (ROS) was defined as sampling with replacement from boundary regions. *Random under-sampling* (RUS) was defined as sampling with replacement from non-boundary regions. Lastly, we considered *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote].

A total of 78 cell line specific functional genomic annotations were used to build the feature space of downstream random forest models (Supplementary Table 2). The annotations could be broken down into three sets including histone modifications (HM), cis-regulatory elements (BroadHMM), and transcription factor binding sites (TFBS). Four different feature engineering procedures were developed for each genomic element (Figure 1 B). *Peak Signal* was defined as the [???include definition]. Counts of overlaps (*Overlap Counts*, OC) were defined as the total number of genomic annotation regions that overlapped within a given genomic bin. Percents of overlaps (*Overlap Percent*, OP) were defined as the proportion of a bin overlapped by an annotation region. *Distance* was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations. A log2-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). These feature engineering procedures were performed to improve boundary predictions by accounting for direct (OC), proportional (Signal/OP), and spatial (distance) relationships between TAD boundaries and genomic annotations.

@@

Thus, under our framework, we considered combinations of data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, with a specific engineered feature space *P = {Signal, OC, OP, Distance}*, and re-sampling technique *S = {None, RUS, ROS, SMOTE}* (Figure 2). Once the model inputs were established, a random forest classifier was trained on chromosome-specific genomic bins (excluding chromosome 9), *C = {CHR1, ..., CHR22}*, using 5-fold cross-validation. Model inputs were evaluated by assessing the balanced accuracy (BA) of each combination. <!--The results presented are for the ARROWHEAD boundaries, with PEAKACHU boundaries presented in the Supplementary material. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |S| \times |C|$ different ensemble systems, totaling $2 \times 5 \times 4 \times 4 \times 21 (= 3360)$ models.-->

@@

Using ARROWHEAD, we obtained a total of 8052, 7676, 4670, 2349, and 1031 called TADs  across each resolution respectively (Table 1). From the called TADs, a total of 15468, 14253, 8363, 4224, and 1883 unique TAD boundaries were obtained (Table 1). The resolution-specific approach we developed to create boundary vs non-boundary regions resulted in a total of 535363, 267682, 107073, 53537, and 26768 genomic bins. The class imbalance among the genomic bins was both severe and invariant toward resolution, resulting in 0.03, 0.05, 0.08, 0.08, and 0.07 for each resolution respectively (Table 1). Boundary data summaries were similar for K562 cell line, as well as for PEAKACHU defined boundaries (Supplementary Tables 3-4).

@@

Expectedly, when using data with class imbalance present, that is no re-sampling, the models exhibited low balanced accuracies, with very little variability among different resolutions (Figure 3). Similarly poor performances were found when using ROS as well. However, RUS and SMOTE re-sampling led to drastic improvement in performance, especially at higher resolutions. We found that RUS narrowly out performed SMOTE as the most optimal class balancing technique for all resolutions and predictor types. These results suggest that simple randomly under-sampling alone can drastically improve TAD boundary prediction. Additionally, we found that at higher resolutions, random forest classifiers built using a distance-type feature space yielded substantially higher MCCs when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with comparing class balancing techniques, this improvement was less evident at lower resolutions. These results were found to be consistent for the K562 cell line (Supplementary Figure 2). Contrary to ARROWHEAD defined boundaries, random forest models built on PEAKACHU loop boundaries exhibited optimal performance when implemented on SMOTE re-sampled data for both GM12878 and K562 cell lines (Supplementary Figure 3). In summary, our computational framework showed that using random under-sampling, distance-type predictors, and Hi-C data at 5kb resolution yields an optimally performing cell line-specific random forest classification model for TAD boundary region prediction.

@@

We then aimed to compare the performances of genomic class-specifc models. In doing so, we evaluated performances for an optimally built random forest model using only histone modifications (HM), cis-regulatory elements (BroadHMM), and transcription factor binding sites (TFBS) separately. We found that models built solely on TFBS outperformed other class-specific models suggesting that TFBS are the primary drivers of TAD and loop boundary formation in both the GM12878 and K562 cell lines (Figure 4A; Supplementary Figures 4A,5A,6A).

@@

We sought to further optimize our TAD prediction model by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all chromosomes. Under our framework, we were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B,5B,6B). However, given that we trained our model on chromosome-specific data, the most significant annotations were found to vary for each chromosome. To determine the most common TFBS, we clustered the predictive importances across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD boundary regions for all chromosomes (Figure 4C; Supplementary Figures 4C,5C,6C). Therefore, our framework was able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@fudenberg2016formation; @alipour2012self]. We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance.

>>>>

Cell type-specific genomic annotations were used to build the predictor space for our TAD boundary prediction framework. Bins were annotated by the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average [??? Is that correct? Or, max?] signal strength of the corresponding annotation (*Peak Signal Strength*), and the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*). A log2-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1).

@@

We developed a computational framework<!--, refered to as *TADRF*,--> aimed at determining the optimal set of data level characteristics necessary for topologically associated domain (TAD) boundary prediction. Random forest machine learning algorithm was selected as the most robust to overfitting [??? Read 10.1002/widm.1072 and 10.1002/widm.1301, elaborate, cite] We then leveraged our optimally built random forest to precisely predict the genomic coordinates of both TAD and loop boundaries at base pair resolution.

Our framework is resolution-specific and uses arbitrary called TAD and loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). TADs were subsequently called [??? Did you call them?] using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries we opted to use boundaries obtained by the PEAKACHU tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results]

Boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a boundary ($Y = 0$) (Figure 1A, see Methods). The total number of TADs, TAD boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 3). The number of non-boundary regions highly outnumbered boundary regions, and these results were consistent for loop boundaries (Supplementary Table 4). Such disproportional presence of examples in one class is known as "class imbalance" problem that negatively affects model performance [??? Add refs]. To address class imbalance, we evaluated the effect of three re-sampling techniques on prediction performance. *Random over-sampling* (ROS) was defined as sampling with replacement from boundary regions. *Random under-sampling* (RUS) was defined as sampling with replacement from non-boundary regions. Lastly, we considered *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote] (see Methods). These strategies allowed us to select the best performing model.

A total of 78 cell line specific functional genomic annotations were used to build the feature space of downstream random forest models. These annotations included BroadHMM shromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS, Supplementary Table 2). Four feature engineering procedures were developed for each genomic annotation (Figure 1B, see Methods) with the goal to improve boundary predictions by accounting for signal strength association (Signal), direct (OC), proportional (OP), and spatial (Distance) relationships between TAD boundaries and genomic annotations.

@@

In summary, we considered combinations of data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, with a specific engineered feature space *P = {Signal, OC, OP, Distance}*, and re-sampling technique *S = {None, RUS, ROS, SMOTE}* (Figure 2). Once the model inputs were established, a random forest classifier was trained on chromosome-specific genomic bins (excluding chromosome 9), *C = {CHR1, ..., CHR22}*, using 5-fold cross-validation. Model inputs were evaluated by assessing the balanced accuracy (BA) of each combination. <!--The results presented are for the ARROWHEAD boundaries, with PEAKACHU boundaries presented in the Supplementary material. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |S| \times |C|$ different ensemble systems, totaling $2 \times 5 \times 4 \times 4 \times 21 (= 3360)$ models.--> [??? When we have all results, Peakachu results will likely need to be described in parallel. They are about loops, theoretically different from TAD boundaries. For discussion]

@@

<!-- Using ARROWHEAD, we obtained a total of 8052, 7676, 4670, 2349, and 1031 called TADs  across each resolution respectively (Table 1). From the called TADs, a total of 15468, 14253, 8363, 4224, and 1883 unique TAD boundaries were obtained (Table 1). The resolution-specific approach we developed to create boundary vs non-boundary regions resulted in a total of 535363, 267682, 107073, 53537, and 26768 genomic bins. The class imbalance among the genomic bins was both severe and invariant toward resolution, resulting in 0.03, 0.05, 0.08, 0.08, and 0.07 for each resolution respectively (Table 1). Boundary data summaries were similar for K562 cell line, as well as for PEAKACHU defined boundaries (Supplementary Tables 3-4). -->

@@

Expectedly, when using data with class imbalance present, that is, the data were used as-is, the models exhibited low balanced accuracies, with very little variability among different resolutions (Figure 3). Similarly, poor performances were found when using ROS as well. However, RUS and SMOTE re-sampling led to drastic improvement in performance, especially at higher resolutions. We found that RUS marginally outperformed SMOTE as the most optimal class balancing technique for all resolutions and predictor types. These results suggest that simple randomly under-sampling alone can drastically improve TAD boundary prediction.

Additionally, we found that at higher resolutions, random forest classifiers built using a distance-type feature space yielded substantially higher balanced accuracies when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with comparing class balancing techniques, this improvement was less evident at lower resolutions. These results were found to be consistent for the K562 cell line (Supplementary Figure 2). Random forest models built on PEAKACHU loop boundaries with SMOTE resampling marginally outperformed RUS for both GM12878 and K562 cell lines (Supplementary Figure 3). In summary, random under-sampling and SMOTE, distance-type predictors, and Hi-C data at 5kb resolution yields an optimally performing cell line-specific random forest classification model for TAD and loop boundary predictions.

@@

We hypothesized that the type of genomic annotations may affect the predictive performance. Using the aforementioned optimal settings, we built separate random forest models using BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, and these results were consistent for loop boundaries (Figure 4A; Supplementary Figures 4A,5A,6A). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both the GM12878 and K562 cell lines.

@@

We sought to further optimize our TAD prediction model by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B,5B,6B). However, given that we trained our model on chromosome-specific data, the most significant annotations were found to vary for each chromosome. To determine transcription factors most important for boundary predigtion, we clustered the predictive importances of top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD and loop boundaries for all chromosomes (Figure 4C; Supplementary Figures 4C,5C,6C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our framework was able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@fudenberg2016formation; @alipour2012self].
